---
title: "class11"
author: "Amit Subramanian"
format: pdf
---


## Section 1. Proportion of G|G in MXL population


Let's read in our csv file we downloaded from Ensemble.

```{r}
mxl <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100
```

Let's take a look at another population, for example: GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G of GBR

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100)
```

We now know that the variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.


## Section 4: Population Scale Analysis


One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale.

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.


> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
# This will tell us how many samples we have

expr <- read.table(url("https://bioboot.github.io/bimm143_S24/class-material/rs8067378_ENSG00000172057.6.txt"))
head(expr)
```

```{r}
library(dplyr)
expr %>%
  group_by(geno) %>%
  summarize(
    sample_size = n (),
    median_expression = median (exp, na.rm=TRUE)
  )
```

```{r}
nrow(expr)
```

We need to find how many there are of each type using 'table()'.

```{r}
table(expr$geno)
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Let's create a boxplot

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) +
  aes(x = geno, y = exp, fill = geno)+
  geom_boxplot(notch = TRUE)

```

From this plot we are able to gather that the relative expression level of A|A is higher than the relative expression level of G|G.  Additionally, the SNP does appear to affect the expression of ORMDL3.
